\Rpackage{MSnID} Package for Handling MS/MS Identifications

### 2014-04-02 inst/doc/msnid_vignette.Rnw
tweaked vignette

\subsection{Parent ion mass measurement accuracy}
Assuming both \Rcode{calculatedMassToCharge} and \Rcode{experimentalMassToCharge} are present in \Rcode{names(msnid)}, one can access parent ion mass measurement in points per million (ppm) units.
<<label=ppmOriginal, fig=TRUE, include=FALSE, width=9>>=
ppm <- mass_measurement_error(msnid)
ggplot(as.data.frame(ppm), aes(x=ppm)) + geom_histogram(binwidth=100)
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-ppmOriginal}
\end{center}
\pagebreak[0]
Note, although the MS/MS search was done with +/- 20ppm parent ion mass tolerance, error stretch over 1000 in ppm units. The reason is that setting of the MS/MS search engine MS-GF+ assumed imperfect peak picking for fragmentation and considered peptides that were +/- 1 Da (\textsuperscript{13}C-\textsuperscript{12}C to be exact) off.
<<label=deltaMass, fig=TRUE, include=FALSE, width=9>>=
dM <- with(psms(msnid), (experimentalMassToCharge-calculatedMassToCharge)*chargeState)
x <- data.frame(dM, isDecoy=msnid$isDecoy)
ggplot(x, aes(x=dM, fill=isDecoy)) + geom_histogram(position='stack')
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-deltaMass}
\end{center}
\pagebreak[0]
Selection and reporting of wrong monoisotopic masses can be corrected.
<<label=ppmCorrectedMass, fig=TRUE, include=FALSE, width=9>>=
msnid.fixed <- correct_peak_selection(msnid)
ppm <- mass_measurement_error(msnid.fixed)
ggplot(as.data.frame(ppm), aes(x=ppm)) + geom_histogram(binwidth=0.25)
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-ppmCorrectedMass}
\end{center}
\pagebreak[0]
Alternatively, one can simply apply a filter to remove any matches that do not fit the +/- 20 ppm tolerance.
<<label=ppmFiltered20, fig=TRUE, include=FALSE, width=9>>=
msnid.chopped <- apply_filter(msnid, "abs(mass_measurement_error(msnid)) < 20")
ppm <- mass_measurement_error(msnid.chopped)
ggplot(as.data.frame(ppm), aes(x=ppm)) + geom_histogram(binwidth=0.25)
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-ppmFiltered20}
\end{center}
\pagebreak[0]
!NOTE! Make sure the text is correct.
For further processing we'll consider \Rcode{msnid.chopped} data that is adjusted for 1 Da errors in picking of wrong monoisotopic peak. Note, if the center of the histogram is significantly shifted from zero, \Rcode{experimentalMassToCharge} can be recalibrated.
<<label=ppmRecalibrated, fig=TRUE, include=FALSE, width=9>>=
msnid <- recalibrate(msnid.chopped)
ppm <- mass_measurement_error(msnid)
ggplot(as.data.frame(ppm), aes(x=ppm)) + geom_histogram(binwidth=0.25)
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-ppmRecalibrated}
\end{center}
\pagebreak[0]

\subsection{Defining and optimizing filters for MS/MS identifications}
First, let's define criteria we are going to use for data filtration. First will be based on MS-GF+ Spectrum E-value, second will be absolute mass measurement error of the parent ion.
<<label=msmsScoreDistribution, fig=TRUE, include=FALSE, width=9>>=
msnid$msmsScore <- -log10(msnid$`ms-gf:specevalue`)
params <- psms(msnid)[,c("msmsScore","isDecoy")]
ggplot(params) + geom_density(aes(x = msmsScore, color = isDecoy, ..count..))
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-msmsScoreDistribution}
\end{center}
\pagebreak[0]
<<label=absPpmDistribution, fig=TRUE, include=FALSE, width=9>>=
msnid$absParentMassErrorPPM <- abs(mass_measurement_error(msnid))
params <- psms(msnid)[,c("absParentMassErrorPPM","isDecoy")]
ggplot(params) + geom_density(aes(x = absParentMassErrorPPM, color = isDecoy, ..count..))
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-absPpmDistribution}
\end{center}
\pagebreak[0]
MS/MS fiters are handled by a special \Rclass{MSnIDFilter} class objects. Individual filtering criteria can be set by name (that is present in \Rcode{names(msnid)}), comparison operator (>, <, = , ...) defining what to retain, and a threshold value.
<<>>=
filtObj <- MSnIDFilter(msnid)
filtObj$absParentMassErrorPPM <- list(comparison="<", threshold=10.0)
filtObj$msmsScore <- list(comparison=">", threshold=7.0)
# print filter
show(filtObj)
# evaluate filter
evaluate_filter(msnid, filtObj, level="PSM")
evaluate_filter(msnid, filtObj, level="Peptide")
evaluate_filter(msnid, filtObj, level="Accession")
@
The threshold values in the example above are not necessarily optimal and set just be in the range of probable values. Filters can be optimized to ensure maximum number of identifications (peptide-to-spectrum matches, unique peptide sequences or proteins) within given FDR upper limit.
<<>>=
# brute-force optimization by enumeration all the parameter combinations
# these should be good starting parameters for follow-up optimizations
system.time({
   filtObj.grid <- optimize_filter(filtObj, msnid, fdr.max=0.01, 
                                  method="Grid", level="Peptide", n.iter=500)})
show(filtObj.grid)
# (absParentMassErrorPPM < 2) & (msmsScore > 7.8)

# Nelder-Mead optimization
set.seed(0)
system.time({
   filtObj.nm <- optimize_filter(filtObj.grid, msnid, fdr.max=0.01, 
                          method="Nelder-Mead", level="Peptide", n.iter=500)})
show(filtObj.nm)
# (absParentMassErrorPPM < 3) & (msmsScore > 7.8)

# simulated annealing optimization
set.seed(0)
system.time({
   filtObj.sann <- optimize_filter(filtObj.grid, msnid, fdr.max=0.01, 
                                  method="SANN", level="Peptide", n.iter=500)})
show(filtObj.sann)
# (absParentMassErrorPPM < 2.2) & (msmsScore > 7.6)
@
Let's check which filter turned out to be the best. The differece between the optimized filters is likely to be very small.
<<>>=
evaluate_filter(msnid, filtObj, level="Peptide")
evaluate_filter(msnid, filtObj.grid, level="Peptide")
evaluate_filter(msnid, filtObj.nm, level="Peptide")
evaluate_filter(msnid, filtObj.sann, level="Peptide")
@
We'll choose and apply filter produced by simulated annealing optimization procedure. Identifications that matched to decoy and contaminant protein sequences can be removed by providing filters in the forms of text strings that will be evaluated in the context of PSM table.
<<>>=
# apply the optimized filter
msnid <- apply_filter(msnid, filtObj.sann)
show(msnid)
# let's remove reverse/decoy and Contaminants
msnid <- apply_filter(msnid, "!isDecoy")
show(msnid)
msnid <- apply_filter(msnid, "!grepl('Contaminant',Accession)")
show(msnid)
@

@@

\pagebreak[4]

>>>>

\subsection{Parent ion mass measurement accuracy}
Assuming both \Rcode{calculatedMassToCharge} and \Rcode{experimentalMassToCharge} are present in \Rcode{names(msnid)}, one can access parent ion mass measurement in points per million (ppm) units.
<<label=ppmOriginal, fig=TRUE, include=FALSE, width=9>>=
ppm <- mass_measurement_error(msnid)
ggplot(as.data.frame(ppm), aes(x=ppm)) + geom_histogram(binwidth=100)
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-ppmOriginal}
\end{center}
\pagebreak[0]
Note, although the MS/MS search was done with +/- 20ppm parent ion mass tolerance, error stretch over 1000 in ppm units. The reason is that setting of the MS/MS search engine MS-GF+ assumed imperfect peak picking for fragmentation and considered peptides that were +/- 1 Da (\textsuperscript{13}C-\textsuperscript{12}C to be exact) off.
<<label=deltaMass, fig=TRUE, include=FALSE, width=9>>=
dM <- with(psms(msnid), (experimentalMassToCharge-calculatedMassToCharge)*chargeState)
x <- data.frame(dM, isDecoy=msnid$isDecoy)
ggplot(x, aes(x=dM, fill=isDecoy)) + 
   geom_histogram(position='stack', binwidth=0.1)
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-deltaMass}
\end{center}
\pagebreak[0]
Selection and reporting of wrong monoisotopic masses can be corrected.
<<label=ppmCorrectedMass, fig=TRUE, include=FALSE, width=9>>=
msnid.fixed <- correct_peak_selection(msnid)
ppm <- mass_measurement_error(msnid.fixed)
ggplot(as.data.frame(ppm), aes(x=ppm)) + geom_histogram(binwidth=0.25)
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-ppmCorrectedMass}
\end{center}
\pagebreak[0]
Alternatively, one can simply apply a filter to remove any matches that do not fit the +/- 20 ppm tolerance.
<<label=ppmFiltered20, fig=TRUE, include=FALSE, width=9>>=
msnid.chopped <- apply_filter(msnid, "abs(mass_measurement_error(msnid)) < 20")
ppm <- mass_measurement_error(msnid.chopped)
ggplot(as.data.frame(ppm), aes(x=ppm)) + geom_histogram(binwidth=0.25)
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-ppmFiltered20}
\end{center}
\pagebreak[0]
!NOTE! Make sure the text is correct.
For further processing we'll consider \Rcode{msnid.chopped} data that is adjusted for 1 Da errors in picking of wrong monoisotopic peak. Note, if the center of the histogram is significantly shifted from zero, \Rcode{experimentalMassToCharge} can be recalibrated.
<<label=ppmRecalibrated, fig=TRUE, include=FALSE, width=9>>=
msnid <- recalibrate(msnid.chopped)
ppm <- mass_measurement_error(msnid)
ggplot(as.data.frame(ppm), aes(x=ppm)) + geom_histogram(binwidth=0.25)
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-ppmRecalibrated}
\end{center}
\pagebreak[0]

\subsection{Defining and optimizing filters for MS/MS identifications}
First, let's define criteria we are going to use for data filtration. First will be based on MS-GF+ Spectrum E-value, second will be absolute mass measurement error of the parent ion.
<<label=msmsScoreDistribution, fig=TRUE, include=FALSE, width=9>>=
msnid$msmsScore <- -log10(msnid$`ms-gf:specevalue`)
params <- psms(msnid)[,c("msmsScore","isDecoy")]
ggplot(params) + geom_density(aes(x = msmsScore, color = isDecoy, ..count..))
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-msmsScoreDistribution}
\end{center}
\pagebreak[0]
<<label=absPpmDistribution, fig=TRUE, include=FALSE, width=9>>=
msnid$absParentMassErrorPPM <- abs(mass_measurement_error(msnid))
params <- psms(msnid)[,c("absParentMassErrorPPM","isDecoy")]
ggplot(params) + geom_density(aes(x = absParentMassErrorPPM, color = isDecoy, ..count..))
@
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics-absPpmDistribution}
\end{center}
\pagebreak[0]
MS/MS fiters are handled by a special \Rclass{MSnIDFilter} class objects. Individual filtering criteria can be set by name (that is present in \Rcode{names(msnid)}), comparison operator (>, <, = , ...) defining what to retain, and a threshold value.
<<>>=
filtObj <- MSnIDFilter(msnid)
filtObj$absParentMassErrorPPM <- list(comparison="<", threshold=10.0)
filtObj$msmsScore <- list(comparison=">", threshold=10.0)
# print filter
show(filtObj)
# evaluate filter
evaluate_filter(msnid, filtObj, level="PSM")
evaluate_filter(msnid, filtObj, level="Peptide")
evaluate_filter(msnid, filtObj, level="Accession")
@
The threshold values in the example above are not necessarily optimal and set just be in the range of probable values. Filters can be optimized to ensure maximum number of identifications (peptide-to-spectrum matches, unique peptide sequences or proteins) within given FDR upper limit.
<<>>=
# brute-force optimization by enumeration all the parameter combinations
# these should be good starting parameters for follow-up optimizations
system.time({
   filtObj.grid <- optimize_filter(filtObj, msnid, fdr.max=0.01, 
                                  method="Grid", level="Peptide", n.iter=500)})
show(filtObj.grid)
# (absParentMassErrorPPM < 2) & (msmsScore > 7.8)

# Fine tuning. Nelder-Mead optimization
system.time({
   filtObj.nm <- optimize_filter(filtObj.grid, msnid, fdr.max=0.01, 
                          method="Nelder-Mead", level="Peptide", n.iter=500)})
show(filtObj.nm)
# (absParentMassErrorPPM < 3) & (msmsScore > 7.8)
@
Evaluating filter based on a good guess and optimization. Finally we'll apply
optimized filter to proceed with further steps in the pipeline.
<<>>=
# comparing orignal and optimized filters
evaluate_filter(msnid, filtObj, level="Peptide")
evaluate_filter(msnid, filtObj.nm, level="Peptide")
# apply the optimized filter
msnid <- apply_filter(msnid, filtObj.nm)
show(msnid)
@
Identifications that matched to decoy and contaminant protein sequences can be removed by providing filters in the forms of text strings that will be evaluated in the context of PSM table.
<<>>=
# let's remove reverse/decoy and Contaminants
msnid <- apply_filter(msnid, "!isDecoy")
show(msnid)
msnid <- apply_filter(msnid, "!grepl('Contaminant',Accession)")
show(msnid)
@

@@
